---
title: "EDA"
output:
  html_document: default
  pdf_document: default
date: "2024-05-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Exploratory Data Analysis of UCIs "Apartment for Rent Classified"

## **1. Data Loading**

```{r}
# Set the working directory to the folder containing your data
setwd("C:/code/Courses/DataAnalytics/Project/RentAnalytics")
```
```{r}
df <- read.csv('apartment_data_final.csv')
```

## **2. Initial Data Exploration**

```{r}
data <- df
```

```{r}
dim(data)
```
#### Data Summary

```{r}
str(data)
```
#### Calculating the percentage of NA values in each column and sorting them in descending

```{r}
sort(colMeans(is.na(data)), decreasing = TRUE)

```
### Identifying Variables with Missing Values

```{r}
vars_with_na <- names(data)[colSums(is.na(data)) > 0]
```


```{r}
na_percentage <- colMeans(is.na(data[vars_with_na]))
```

```{r}
na_percentage
```
#### Aggregating Data for Analysis

This analysis helps understand regional price differences.

```{r}
library(dplyr)

# Calculating the average price by region and sorting the results
average_price_by_region <- data %>%
  group_by(us_region) %>%
  summarise(average_price = mean(price, na.rm = TRUE)) %>%
  arrange(desc(average_price))

# Printing the results
print(average_price_by_region)
```


```{r}
library(dplyr)

# Average price by state
state_avg_price <- df %>%
  group_by(state) %>%
  summarise(average_price = mean(price, na.rm = TRUE))
print(state_avg_price)

# Average price by presence of photos
photo_avg_price <- df %>%
  group_by(has_photo) %>%
  summarise(average_price = mean(price, na.rm = TRUE))
print(photo_avg_price)

# Average price by region and division
region_division_avg_price <- df %>%
  group_by(us_region, us_division) %>%
  summarise(average_price = mean(price, na.rm = TRUE))
print(region_division_avg_price)

# Average price for studios
studio_avg_price <- df %>%
  group_by(studio) %>%
  summarise(average_price = mean(price, na.rm = TRUE))
print(studio_avg_price)

# Average price considering pet policy
pet_avg_price <- df %>%
  group_by(dogs_allowed, cats_allowed) %>%
  summarise(average_price = mean(price, na.rm = TRUE))
print(pet_avg_price)
```

- **Group by state and calculate average price**  
This would allow you to see the average price of properties in each state.

- **Group by whether the property has photos and calculate average price**
This would help you understand if having photos affects the price.

- **Group by US region and division and calculate average price**
How price varies across different regions and divisions of the US.

- **Group by two categories—dogs_allowed and cats_allowed and calculate average price**
Helps to understand the impact of pet-friendliness on rental costs.


#### Exploring Boolean and Numerical Variables


```{r}
bool_vars <- names(data)[sapply(data, function(x) length(unique(x)) == 2)]

# Display the first few rows of these columns
head(data[bool_vars])
```

```{r}
# Identify numerical variables, excluding boolean ones
num_vars <- names(data)[sapply(data, is.numeric) & !names(data) %in% bool_vars]

print(paste('Number of numerical variables: ', length(num_vars)))
head(data[num_vars])
```

#### Geospatial Visualization

Map of the world and plots the geospatial data on it, visualizing the locations of the apartments on a global map. This helps understand the geographical distribution of the dataset's entries.

```{r}
library(sf)
library(ggplot2)

# Create an sf object (adjust as per your previous transformations)
data_sf <- st_as_sf(data, coords = c("longitude", "latitude"), crs = 4326, agr = "constant")

# Read the world map (adjust the file path to your specific file)
world <- st_read(system.file("shape/nc.shp", package="sf"))  # Update with your path

# Plotting with adjustments
ggplot() +
  geom_sf(data = world, fill = "gray90") +  # Adjusting world map color
  geom_sf(data = data_sf, aes(color = 'red'), shape = 19, size = 2, alpha = 0.5) +  # Smaller, transparent red points
  theme_minimal() +
  labs(title = "Geospatial Plot", x = "Longitude", y = "Latitude")

```


#### Identifying Discrete Variables

- Identifies discrete numerical variables in the dataset, defined here as those numerical variables that have fewer than 20 unique values and are not 'id' or 'price'.

```{r}
# Load necessary library
library(dplyr)

# Identify discrete variables from the set of numerical variables, excluding 'id' and 'price'
discrete_vars <- num_vars[num_vars != "id" & num_vars != "price" & sapply(data[num_vars], function(x) length(unique(x)) < 20)]

# Print the number of discrete variables
print(paste('Number of discrete variables: ', length(discrete_vars)))

# Display the first few rows of these discrete variables
head(data[discrete_vars])
```


#### Visualizing Discrete Variables

```{r}
library(ggplot2)
library(dplyr)

analyse_discrete <- function(df, var) {
  # Create a variable symbol from the string
  var_sym <- rlang::sym(var)

  # Creating a summary of median prices by the discrete variable
  grs <- df %>%
    group_by(!!var_sym) %>%
    summarise(price_median = median(price, na.rm = TRUE)) %>%
    ungroup()

  # Plotting the results
  p <- ggplot(grs, aes(x = !!var_sym, y = price_median)) +
    geom_bar(stat = "identity", fill = "steelblue") +
    theme_minimal() +
    labs(title = toupper(var), x = var, y = "Median Price") +
    theme(plot.title = element_text(hjust = 0.5)) # Center the plot title
  
  print(p)
}
```

- This function visualizes the relationship between each discrete variable and the 'price' using a bar plot, showing the median price for each category of the discrete variable.

```{r}
# Apply the function to each discrete variable
for (var in discrete_vars) {
  analyse_discrete(data, var)
}

```


#### Identifying and Analyzing Continuous Variable Distributions

- Identifies continuous variables, considered as those not already classified as discrete and excluding 'id'.

```{r}
cont_vars <- num_vars[!(num_vars %in% c(discrete_vars, "id"))]

print(paste('Number of continuous variables: ', length(cont_vars)))
head(data[cont_vars])
```


```{r}
library(ggplot2)
library(dplyr)
library(moments)

analyse_continuous <- function(df, var) {
  # Ensure the variable is a symbol for tidy evaluation
  var_sym <- rlang::sym(var)

  # Remove NA values and prepare data
  df <- df %>% 
    filter(!is.na(!!var_sym))

  # Create the distribution plot
  p <- ggplot(df, aes(x = !!var_sym)) +
    geom_histogram(aes(y = ..density..), bins = 30, fill = "blue", alpha = 0.7) +
    geom_density(color = "red", size = 1.5) +
    labs(title = var, x = var, y = "Density") +
    theme_minimal()

  # Calculate skewness and kurtosis
  skewness <- moments::skewness(df[[var]])
  kurtosis <- moments::kurtosis(df[[var]])

  # Add annotations for skewness and kurtosis
  p <- p + annotate("text", x = Inf, y = Inf, label = sprintf("Skewness=%.2f Kurtosis=%.2f", skewness, kurtosis), 
                    hjust = 1.1, vjust = 2, size = 5, color = "black")

  # Print the plot
  print(p)
}
```

- This function visualizes the distribution of each continuous variable using histograms, including metrics like skewness and kurtosis for deeper insights into each distribution's shape.

```{r}
# Apply the function to each continuous variable
for (var in cont_vars) {
  analyse_continuous(data, var)
}
```

#### Logarithmic Transformation on Continuous Variables

```{r}
library(ggplot2)
library(dplyr)
library(moments)  # For skewness and kurtosis

analyse_transformed_continuous <- function(df, var) {
  # Ensure the variable is a symbol for tidy evaluation
  var_sym <- rlang::sym(var)
  
  # Remove NA values
  df <- df %>%
    filter(!is.na(!!var_sym))

  # Skip transformation for 'latitude' or 'longitude'
  if (var %in% c('latitude', 'longitude')) {
    message(paste("Skipping transformation for", var))
  } else {
    # Apply logarithmic transformation with +1 to handle zero and negative values
    df <- df %>%
      mutate(!!var_sym := log1p(!!var_sym))
  }

  # Create the distribution plot
  p <- ggplot(df, aes(x = !!var_sym)) +
    geom_histogram(aes(y = ..density..), bins = 30, fill = "blue", alpha = 0.7) +
    geom_density(color = "red", size = 1.5) +
    labs(title = var, x = paste("Transformed", var), y = "Density") +
    theme_minimal()

  # Calculate skewness and kurtosis
  skewness_val <- skewness(df[[var]], na.rm = TRUE)
  kurtosis_val <- kurtosis(df[[var]], na.rm = TRUE) - 3  # Adjust kurtosis to match Python's definition

  # Add annotations for skewness and kurtosis
  p <- p + annotate("text", x = Inf, y = Inf, label = sprintf("Skewness=%.2f Kurtosis=%.2f", skewness_val, kurtosis_val), 
                    hjust = 1.1, vjust = 2, size = 5, color = "black")

  # Print the plot
  print(p)
}
```

- Applies a logarithmic transformation to each continuous variable (except for geographical coordinates like latitude and longitude) and visualizes their new distributions to often normalize data and reduce skewness.

```{r}
# Apply the function to each continuous variable
for (var in cont_vars) {
  analyse_transformed_continuous(data, var)
}
```


#### Visualizing Outliers in Continuous Variables
```{r}
# Removing rows where the 'price' column has NA values
df <- na.omit(df, cols = "price")
```

```{r}
library(ggplot2)
library(dplyr)

find_outliers <- function(df, var) {
  # Ensure the variable is a symbol for tidy evaluation
  var_sym <- rlang::sym(var)
  
  # Skip transformation for 'latitude' or 'longitude'
  if (var %in% c('latitude', 'longitude')) {
    message(paste("Skipping", var))
  } else {
    # Apply logarithmic transformation to handle zero and negative values
    df <- df %>%
      mutate(!!var_sym := log1p(!!var_sym))
  }
  
  # Plotting the boxplot
  p <- ggplot(df, aes_string(x = "1", y = as.character(var_sym))) +
    geom_boxplot() +
    labs(title = var, y = paste("Transformed", var)) +
    theme_minimal() +
    theme(axis.text.x=element_blank(), axis.ticks.x=element_blank())  # Hide x-axis details
  
  # Print the plot
  print(p)
}
```

```{r}
# Apply the function to each continuous variable
for (var in cont_vars) {
  find_outliers(data, var)
}
```

#### Outlier Detection


```{r}
library(ggplot2)

# Create the scatter plot
p <- ggplot(data, aes(x = square_feet, y = price)) +
  geom_point() +  # Add points
  labs(x = "square_feet", y = "price", title = "Price vs. Square Feet") +
  theme_minimal() +  # Use a minimal theme
  theme(axis.title.x = element_text(size = 13),  # Customize font size for x label
        axis.title.y = element_text(size = 13))  # Customize font size for y label

# Display the plot
print(p)
```


```{r}
library(dplyr)

out_iqr <- function(df, column) {
  # Calculate the IQR
  q25 <- quantile(df[[column]], 0.25, na.rm = TRUE)
  q75 <- quantile(df[[column]], 0.75, na.rm = TRUE)
  iqr <- q75 - q25
  
  # Calculate the outlier cutoff
  cut_off <- iqr * 1.5
  lower <- q25 - cut_off
  upper <- q75 + cut_off
  
  # Output the IQR and bounds
  print(paste("The IQR is", iqr))
  print(paste("The lower bound value is", lower))
  print(paste("The upper bound value is", upper))
  
  # Calculate the number of outliers
  num_outliers <- sum(df[[column]] < lower | df[[column]] > upper, na.rm = TRUE)
  
  return(print(paste("Total number of outliers are", num_outliers)))
}
```

- Calculatethe IQR details for 'price'  

```{r}

out_iqr(df, 'price')
```

- Calculate the IQR details for 'bedrooms'  

```{r}
out_iqr(df, 'bedrooms')
```

- Calculate he IQR details for 'bathrooms'  

```{r}
out_iqr(df, 'bathrooms')
```
#### Categorical Variable Analysis

Identifies categorical variables and then counts and sorts the unique values in each of these categorical variables to understand their diversity.

```{r}
library(dplyr)

# Identifying categorical variables (assuming 'O' stands for object type in Python)
cat_vars <- names(data)[sapply(data, function(x) is.character(x))]

# Calculating the number of unique values for each categorical variable
num_unique <- sapply(data[cat_vars], function(x) length(unique(x)))

# Sorting the number of unique values in descending order
sorted_unique <- sort(num_unique, decreasing = TRUE)

# Display the sorted values
sorted_unique
```


#### Rare Label Analysis

- To identify and print categories within each categorical variable that appear in less than 1% of the observations

```{r}
library(dplyr)

analyse_rare_labels <- function(df, var, threshold = 0.01) {
  # Calculate the frequency of each category
  freq <- df %>% 
    group_by(!!rlang::sym(var)) %>%
    summarise(Count = n(), .groups = 'drop') %>%
    mutate(Frequency = Count / sum(Count))
  
  # Identify rare labels
  rare_labels <- freq %>% 
    filter(Frequency < threshold)
  
  # Print or return results
  if (nrow(rare_labels) == 0) {
    message(paste("No rare labels found in", var))
  } else {
    print(rare_labels)
  }
  
  return(invisible(rare_labels))
}

```

```{r}
# Assuming cat_vars has been defined as shown previously
for (var in cat_vars) {
  print(analyse_rare_labels(data, var, 0.01))
}
```


#### Geographic Filtering and Visualization

Filters for entries that fall within specific geographic coordinates (likely encompassing the contiguous United States) and a price below $2000.

```{r}
library(dplyr)
library(ggplot2)

# Filter data based on longitude, latitude, and price constraints
df <- data %>%
  filter(longitude > -130, longitude < -60, latitude > 25, latitude < 50, price < 2000)

# Create a scatter plot of longitude vs. latitude
ggplot(df, aes(x = longitude, y = latitude)) +
  geom_point() +  # This adds the scatter plot points
  labs(x = "Longitude", y = "Latitude", title = "Scatter Plot of Locations") +
  theme_minimal()  # Uses a minimal theme for the plot
```

#### Correlation Analysis

Calculates the correlation matrix for numerical columns in the dataset to identify relationships between different numerical variables.

```{r}
library(corrplot)
library(reshape2)

# Create a dataframe containing only numeric columns
numeric_df <- data %>%
  select(where(is.numeric))

# Assuming numeric_df is your DataFrame with only numerical columns
corr_matrix <- cor(numeric_df, use = "complete.obs")  # Computes correlation matrix, handling NA values

# Melt the correlation matrix for ggplot2
melted_corr_matrix <- melt(corr_matrix)

# Plot the heatmap
ggplot(melted_corr_matrix, aes(Var1, Var2, fill = value)) +
  geom_tile() +  # Create tiles for heatmap
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0, limit = c(-1,1), space = "Lab", name="Correlation") +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels for better readability
        axis.title = element_blank())  # Remove axis titles
```